---
title: "DAT537 Final Project"
author: "Adrian Cao"
date: "2022-12-02"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
library(cbw)
library(czzg)
library(ggplot2)
library(gridExtra)
```

## DAT 537 Final Project

Mingxuan(Adrian) Cao BA 2024

\newpage

# Q1

I randomly selected 30 stocks from the S&P 500 companies, there are:

W. R. Berkley Corporation WRB

Expeditors International of Washington Inc. EXPD

Caterpillar Inc. CAT

Quanta Services Inc. PWR

O'Reilly Auto Parts ORLY

Prudential Financial Inc. PRU

Aflac Incorporated AFL

Walmart WMT

Becton Dickinson and Company BDX

Southwest Airlines LUV

NetApp Inc. NTAP

PulteGroup PHM

Verizon Communications Inc. VZ

Weyerhaeuser Company WY

FMC Corporation FMC

Pinnacle West Capital Corporation PNW

Baker Hughes Company Class A BKR

Rollins, Inc. ROL

Regions Financial Corporation RF

Citigroup C

Allstate Corporation ALL

Evergy Inc. EVRG

Illinois Tool Works Inc. ITW

The Estée Lauder Companies Inc. EL

Match Group Inc. MTCH

STERIS Plc STE

FedEx FDX

Wells Fargo & Company WFC

Newmont Corporation NEM

Biogen Inc. BIIB

# Q2

```{r}
datdf = getfinmdat(symbols = c("WRB","EXPD","CAT","PWR","ORLY",
                               "PRU","AFL","WMT","BDX","LUV",
                               "NTAP","PHM","VZ","WY","FMC",
                               "PNW","BKR","ROL","RF","C",
                               "ALL","EVRG","ITW","EL","MTCH",
                               "STE","FDX","WFC","NEM","BIIB"),
                   symnames = c("Berkley","Expeditors","Caterpillar","Quanta","OReilly",
                                "Prudential","Aflac","Walmart","BectonDickinson","Southwest",
                                "NetApp","PulteGroup","Verizon","Weyerhaeuser","FMC",
                                "PinnacleWest","BakerHughes","Rollins","Regions","Citigroup",
                                "Allstate","Evergy","Illinois","EstéeLauder","Match",
                                "STERIS","FedEx","WellsFargo","Newmont","Biogen"),
                   from = as.Date("2004-12-31"),
                   to = as.Date("2018-12-31"))
head(datdf)
```

# Q3

```{r}
data(factor12)
data = factor12
nug = c(4,5,6)

thetamls = mapply(FUN="CZZscant",
                    nu = nug,
                    MoreArgs = list(data = data,mustinclude = "Mkt",m = 2000,nclust = 4),
                    SIMPLIFY = FALSE)
scanord = list()
scanord[[1]] = thetamls[[1]]$scanord
scanord[[2]] = thetamls[[2]]$scanord
scanord[[3]] = thetamls[[3]]$scanord

scanloglik = c()
scanloglik[1] = scanord[[1]][1,]['logmarg']
scanloglik[2] = scanord[[2]][1,]['logmarg']
scanloglik[3] = scanord[[3]][1,]['logmarg']
#nu = 5 is the best out of these three

xbest = names(which(scanord[[which.max(scanloglik)]][1,]==1))
xbest
```

# Q4

```{r}
data = factor12[373:540,]
factor22 = cbind(datdf[,1:30],data)
colnames(factor22)
```
# Q5
```{r}
pout = pricing(xnames = xbest,
               data = factor22)
pout

wp = sum(pout[1:30,3] > .69)
wp
wnp = sum(pout[1:30,3] < -.69)
wnp
```

